Experience Report: Haskell in Computational Biology

### 2012-03-08 mrfy_experience_report.tex
smallverbatim environment, minor edits

\newcommand\figref[1]{Figure~\ref{#1}}

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
For some software, considerations of performance are paramount; this
software is usually written C~or~C++. 
For other software, considerations of convenience, readability, and
productivity are more important;
this software is usually written in a ``scripting'' language like
Perl, Python, Ruby, or~R.\remark{Don't forget audience.}
In~this Experience Report, we show that computational biologists can
successfully replace both these families of languages with Haskell:
\begin{itemize}
\item
Our experimental Haskell programs are as easy to understand and change
 as the programs our group writes using scripting languages.
\item
The performance of our Haskell code is comparable to the performance
of~C++.
\item
We have experienced barriers to entry, of which the most salient has been
the difficulty of understanding and tuning  the performance of Haskell
programs. 
We found the other barriers 
relatively easy to overcome.
\end{itemize}
Our experience is based on MRFy (pronounced ``Murphy''), which
implements a novel family of algorithms for finding remote homologs to 
newly discovered proteins.

@@

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signalling, and more.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
\remark{Lost: there are only 20-odd amino acids}
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence (the \textit{query sequence}), biologists wish to know the cellular
function of the protein that gene codes for.
The best known method of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
Proteins that share structure and function are \textit{homologous}, and thus
the process of identifying similar proteins to a query sequence is called 
\textit{homology detection}.

@@

@@

The protein alignment is used to train a \emph{hidden-Markov model}.
\remark{I've never seen this hypenated, in ANY literature.}
A~hidden-Markov model is a probabilistic finite-state machine which 
can assign a probability to any amino-acid sequence under test.
\remark{What is the biological or computational meaning of the
  probability?}
Proteins whose sequences have higher probabilities are more likely to
be homologous to the proteins in the alignment.
\remark{Should we say ``homologous'' rather than  ``structurally
  related''?
  If~so, where should this jargon be introduced?}
The states of the hidden-Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
If a given column~$j$ of the alignment contains amino acids in a majority
of the protein sequences, it is considered a \textit{consensus column}.
\item
For a given column~$j$ of the alignment, the hidden-Markov model has a
\emph{match state}.
The match state determines the probability that the sequence under
test has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
The match state is more probable if it corresponds to a consensus column.
\item
For every column of the alignment, the hidden-Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the sequence under
test has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$j$, and low probability if there are few gaps in column~$j$.
Thus, the deletion state is more likely in non-consensus columns.
\item
For every column of the alignment, the hidden-Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid that does not correspond to any
consensus column in the alignment.
Thus, the insertion state is more likely in non-consensus columns.
The insertion state assigns high probability to amino acids that are
frequently observed among all proteins, and low probability to amino acids
that are rarely observed.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden-Markov model also has distinguished \emph{begin} and \emph{end} states.
The probability associated with a specific state is conditional upon the previous
state in the model.
Thus, the hidden Markov model can be represented as a directed graph, with each
edge weight corresponding to the probability of a specific state transition.
\figref{plan7} shows a hidden-Markov model derived from an alignment
with just four columns.
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.

@@

>>>>

\newcommand\figref[1]{Figure~\ref{#1}}
\newcommand\secref[1]{Section~\ref{sec:#1}}
\newcommand\seclabel[1]{\label{sec:#1}}

\usepackage{verbatim}

\newenvironment{smallverbatim}{\par\small\verbatim}{\endverbatim}

@@

Computational biologists write software that answers questions about 
sequences of nucleic acids (genomic data) or sequences of amino 
acids (proteomic data). 
For some software, considerations of performance are paramount; this
software is usually written in C~or~C++. 
For other software, considerations of convenience, readability, and
productivity are more important;
this software is usually written in a ``scripting'' language like
Perl, Python, Ruby, or~R.\remark{Don't forget audience.}
In~this Experience Report, we show that computational biologists can
successfully replace both these families of languages with Haskell:
\begin{itemize}
\item
Our experimental Haskell programs are as easy to understand and change
 as the programs our group writes using scripting languages.
\item
The performance of our Haskell code is comparable to the performance
of~C++.
\item
Although we have had to write a new implementation of a standard
algorithm, the cost was low, and the connection between code and
algorithm is unusually strong (\secref{viterbi}).
\item
We have experienced barriers to entry, of which the most salient has been
the difficulty of understanding and tuning  the performance of Haskell
programs. 
We found the other barriers 
relatively easy to overcome.
\end{itemize}
Our experience is based on MRFy (pronounced ``Murphy''), which
implements a novel family of algorithms for finding remote homologs to 
newly discovered proteins.

@@

Proteins are cellular machinery. They interact with one another and with other 
molecules to carry out the functions of living cells: metabolism, regulation, 
signalling, and more.
A~protein's function is determined by its structure, 
and its structure is determined by the sequence of amino acids that
form the protein.
\remark{Lost: there are only 20-odd amino acids}
The amino-acid sequence is ultimately determined by a sequence of
nucleic acids in DNA, which we call a gene.
Given a genetic sequence (the \textit{query sequence}), biologists wish to know the cellular
function of the protein that gene codes for.
\remark{I thought the query sequence was an amino-acid sequence.}
The best known method of discovering such function is
to find other proteins of 
similar structure, which likely share similar function.
Proteins that share structure and function are \textit{homologous}, and thus
the process of identifying similar proteins to a query sequence is called 
\textit{homology detection}.

@@

\seclabel{viterbi}

@@

The protein alignment is used to train a \emph{hidden-Markov model}.
\remark{I've never seen this hypenated, in ANY literature.}
A~hidden-Markov model is a probabilistic finite-state machine which 
can assign a probability to any amino-acid sequence under test.
\remark{What is the biological or computational meaning of the
  probability?}
Proteins whose sequences have higher probabilities are more likely to
be homologous to the proteins in the alignment.
\remark{Should we say ``homologous'' rather than  ``structurally
  related''?
  If~so, where should this jargon be introduced?}
The states of the hidden-Markov model are determined by the alignment:
\remark{I've tried for lots of parallel structure.
Somebody let me know if any of this is right.  ---NR}
\begin{itemize}
\item
If a given column~$j$ of the alignment contains amino acids in a majority
of the protein sequences, it is considered a \textit{consensus column}.
\item
For a given column~$j$ of the alignment, the hidden-Markov model has a
\emph{match state}.
\remark{Time to introduce the elements of Viterbi's equations!}
The match state determines the probability that the sequence under
test has a particular amino acid in column~$j$.
The match state assigns high probability to the amino acids that
appear in column~$j$, and a low but nonzero probability to other amino
acids.
The match state is more probable if it corresponds to a consensus column.
\item
For every column of the alignment, the hidden-Markov model has a
\emph{deletion state}.
The deletion state determines the probability that the sequence under
test has \emph{no} amino acid  in column~$j$, i.e., that it has a gap.
The deletion state has high probability if there are lots of gaps in
column~$j$, and low probability if there are few gaps in column~$j$.
Thus, the deletion state is more likely in non-consensus columns.
\item
For every column of the alignment, the hidden-Markov model has an
\emph{insertion state}.
The insertion state determines the probability that a query sequence
has an ``extra'' amino acid that does not correspond to any
consensus column in the alignment.
Thus, the insertion state is more likely in non-consensus columns.
The insertion state assigns high probability to amino acids that are
frequently observed among all proteins, and low probability to amino acids
that are rarely observed.
\end{itemize}
%%  A position that has residues in a majority of the sequences becomes 
%%  a \textit{match} state in the hidden Markov model; this position may see a 
%%  variety of amino acids due to evolutionary substitution, but it still 
%%  corresponds to a position that a homologous protein should most likely fill.
The hidden-Markov model also has distinguished \emph{begin} and \emph{end} states.
The probability associated with a specific state is conditional upon the previous
state in the model.
Thus, the hidden Markov model can be represented as a directed graph, with each
edge weight corresponding to the probability of a specific state transition.
\figref{plan7} shows a hidden-Markov model derived from an alignment
with just four columns.
Each column in the original alignment corresponds to a \emph{node},
which is a triplet holding an insertion state, a deletion state, and a
match state.

@@

\emph{[This might be a place for comparison with standard
implementations of Viterbi.]}
